A new metric for automatic discovery of periodic patterns in time series

### 2021-03-03 paper/main.Rmd
before application section

In this section, we provide a method to select important harmonies by eliminating all harmonies for which patterns are not significant across x or facet categories through randomization test. Randomization tests (permutation tests) generates a random distribution by re-ordering our observed data and allow to test if the observed data is significantly different from any random distribution. Complete randomness in the measured variable indicates that the process follows a homogeneous underlying distribution over the whole time series, which essentially implies there is no interesting distinction across any different categories of the cyclic granularities.

@@

Typically, a randomization test involves calculating a test statistic, randomly shuffling the data and calculating the test statistic several times to obtain a distribution of the test statistic. We will use this procedure to test if there is any interesting pattern captured by the harmonies, which essentially implies if $wpd_{norm}$ is significantly different from zero. The percentages of times the $wpd_{norm}$ obtained from the permuted data is greater than or equal to the observed $wpd_{norm}$ is the p-value. The randomization test is described as follows:

@@

5.  Steps (3) and (4) are repeated a large number
of times M (M = 200).

@@

8. $95^{th}$ percentile of $wpd^{all}_{sample}$ is computed and stored in $wpd_{threshold}$.

7. If $wpd_{obs_{A, B}}> wpd_{threshold}$, harmony pair $(A, B)$ is selected, otherwise rejected.

@@

Observations are generated from a N(0,1) distribution for each combination of $nx$ and $nfacet$ from the following sets: $nx = nfacet = \{2, 3, 5, 7, 14, 20, 31, 50\}$. For each of the $64$ panels, $ntimes = 500$ observations are drawn for each combination of the categories. The values of $wpd_{norm}$ is obtained for each of the panels for the designs $D_{null}$, $D_{var_x}$, $D_{var_f}$ and $D_{var_{all}}$. $wpd_{threshold}$ is computed from all of these panels together and the number of times a harmony pair $(A, B) \in H_{N_C}$ is selected when in fact it was of the design $D_{null}$ is noted. This entire process is repeated for several null data sets to see the number of times any harmony pair $(A, B) \in H_{N_C}$ is selected under null.

nx = (3, 7, 14)
nfacet = (2, 9, 10)

(3, 2), (7,9), (14, 10) being null sets and others under some other design. Generate 200 repetitions of different data sets with the same design for all the pairs and compute size and power.

@@

Figure \ref{fig:raw} shows the distribution of $wpd$ plotted across different $nx$ and $nfacet$ categories. Since under $D_{null}$, there is no difference in distributions across different categories, we expect the distance measure $wpd$ to reflect that as well and have the same null distribution across categories. But Figure \ref{fig:raw} shows that both the location and scale of the distributions change across panels. This is not desirable under $D_{null}$ as it would mean comparisons of $wpd$ values is not appropriate across different $nx$ and $nfacet$. Figure \ref{fig:quadratic} shows how the median of $wpd$ varies with the total number of distances $nx*nfacet$ for each panel. The median increases abruptly for lower values of $nx*nfacet$ and slowly for higher $nx*nfacet$.

@@

```{r plot-all-new2}
library(patchwork)
(p_null + p_varf)/(p_varx + p_varall)
```

>>>>

In this section, we provide a method to select important harmonies by eliminating all harmonies for which patterns are not significant by employing randomization test. Randomization tests (permutation tests) generates a random distribution by re-ordering our observed data and allow to test if the observed data is significantly different from any random distribution. Complete randomness in the measured variable indicates that the process follows a homogeneous underlying distribution over the whole time series, which essentially implies there is no interesting distinction across any different categories of the cyclic granularities.

@@

A randomization test involves calculating a test statistic, randomly shuffling the data and calculating the test statistic several times to obtain a distribution of the test statistic. But we will use this procedure to obtain a threshold such that harmony pairs with a $wpd_{norm}$ value higher than this threshold will only be considered significant. The process of choosing a threshold is described as follows:

<!-- to test if there is any interesting pattern captured by the harmonies, which essentially implies if $wpd_{norm}$ is significantly different from zero. The percentages of times the $wpd_{norm}$ obtained from the permuted data is greater than or equal to the observed $wpd_{norm}$ is the p-value. The randomization test is described as follows: -->

@@

5.  Steps (3) and (4) are repeated a large number
of times $M$ (M = 200).

@@

8. $99^{th}$ percentiles of $wpd^{all}_{sample}$ is computed and stored in $wpd_{threshold99}$

7. If $wpd_{obs_{A, B}} > wpd_{threshold99}$, harmony pair $(A, B)$ is selected with a 99% level of significance and otherwise rejected.

Similarly, a harmony pair $(A, B)$ is selected with a 95% and 90% level of significance if $wpd_{obs_{A, B}} > wpd_{threshold95}$ and $wpd_{obs_{A, B}} > wpd_{threshold90}$, where  $wpd_{threshold95}$ and $wpd_{threshold90}$ denote the $95^{th}$ and $90^{th}$ percentile of $wpd^{all}_{sample}$ respectively.

@@

Observations are generated from a N(0,1) distribution for each combination of $nx$ and $nfacet$ from the following sets: $nx =  \{3, 7, 14}$ and $nfacet = (2, 9, 10)$. The panel $(3, 2), (7,9), (14, 10)$ are considered to have design $D_{null}$. The panels $(7, 2), (14, 9)$ have design of the form $D_{var_f}$. $(14, 2), (3, 10)$ have design of the form $D_{var_x}$ and the rest are under $D_{var_{null}}$. We generate only one data set for which all these designs were simulated and consider this as the original data set. We generate $200$ repetitions of this experiment with different seeds and compute the proportion of times a panel is rejected when it is under $D_{null}$. We also compute the proportion of times a panel is rejected when it actually belongs to a non-null design. The first proportion is desired to be as small as possible and a higher value of the later is expected. Also, these would consitute to be the estimated size and power of the test.

<!-- 14, 20, 31, 50\}$. For each of the $64$ panels, $ntimes = 500$ observations are drawn for each combination of the categories. The values of $wpd_{norm}$ is obtained for each of the panels for the designs $D_{null}$, $D_{var_x}$, $D_{var_f}$ and $D_{var_{all}}$. $wpd_{threshold}$ is computed from all of these panels together and the number of times a harmony pair $(A, B) \in H_{N_C}$ is selected when in fact it was of the design $D_{null}$ is noted. This entire process is repeated for several null data sets to see the number of times any harmony pair $(A, B) \in H_{N_C}$ is selected under null. -->

<!-- nx = (3, 7, 14) -->
<!-- nfacet = (2, 9, 10) -->

<!-- (3, 2), (7,9), (14, 10) being null sets and others under some other design.  -->

@@

The results for this section is WIP and to be included in details in the Supplementary paper. Also, Figure \ref{fig:plot-all-new2} presents the results of $wpd_{norm}$ from the illustrative designs in Section \ref{sec:computation-wpd}. As expected, the value of $wpd_{norm}$ under null design is the least (a), followed by (b, c and d). Moreover, note the the relative difference in $wpd_{norm}$ values as we move from  a to d, which aligns with the idea of $wpd_{norm}$, since the differences between x categories are up-weighed by design.

@@

```{r plot-all-new2, fig.cap="$wpd_{norm} values for the four illustrative designs are presented here. As expected, the value of wpd_{norm} under null design is the least (a), followed by (b, c and d). Moreover, the relative difference is interesting as c is closer to d than b, since the differences between x categories are up-weighed by design."}
library(patchwork)
(p_null + p_varf)/(p_varx + p_varall)
```
